# Chapter 3 - Logistic regression

<!---
Exploring data and interpreting logistic regresion analysis
-->
```{r echo = FALSE, results='hide', message=FALSE}
library(dplyr)
library(GGally)
library(ggplot2)
```
This time we'll take a look at student alcohol consumption. First off (although I did secretly load some libraries above), let's get the data that I've wrangled into shape earlier:

```{r}
alc <- read.csv("./data/alc_data.csv")
glimpse(alc)
```

Alright, so we have 382 observations of 35 variables, including information on which school they attend, sex, age, urban/rural living etc. Most importantly for our purposes we have information on both workday and weekend alcohol consumption. This I have already averaged to a daily measure of consumption and then arbitrarily decided that if that consumption surpasses the integer 2, the students in question are _high_ consumers of alcohol. In the following examples we shall see the results of this excessive consumption.
This data has been created by joining two separate datasets, one gathered at a math course, the other a Portugese language course. The numeric values have been averaged and rounded while for the non-numeric variables the value from the Math dataset is used. So for example the G1-G3 variables represent rounded averages of students' Math and Portugese school grades.

Now let's get analytical.I've been instructed to _study the relationships between high/low alcohol consumption and some other variables_. So that's what I'll do. There are many variables that have the potential to explain student alcohol consumption and I have to choose 4 of them. I'll pick the obvious ones first: failures, or the number of past class failures (would drive any sane person to the bottle) and sex (guys drink more - duh!). Oh, and age - that has got to be relevant - the older students drink more. Now, my fourth pick might be less obvious though it shouldn't. Internet access at home - if you don't have that, what else you gonna do? Right. So, my variables of choice, in hypothesized order of (absolute) relevance to alcohol consumption:

* sex
* age
* internet
* failures


Here's a quick glance at those variables:
```{r}
summary(alc$sex)
summary(alc$age)
summary(alc$internet)
summary(alc$failures)
```
Ok, so this doesn't tell me so much just yet. About 50-50 female/male. Mean age just under 17 years, span 15-22. Most have a working internet connection at home, and the summary of the failures variable looks really weird.

Let's look at some plots to get a feel for these variables. Let's start with a simple bar plot colored by sex:

```{r}
g1 <- ggplot(data=alc, aes(x = high_use, fill = sex))
g1 + geom_bar()
```

So easy guess, the ratio of non-high to high consumers looks way higher for females. That should be a significant predictor of high alcohol consumption.Now letäs look at age:

```{r}

```