---
title: "Chapter5"
author: "Henrik Husberg"
date: "21.02 2017"
output: html_document
---
# Chapter 5 - Dimensionality reduction techniques

This weeks exercises are all about dimension reductoin. This is very interesting for me because I've been using confirmatory factor analysis in my own research, and I've dabbled a little in PLS Path Modeling on my spare time. So let's see if I can make sense of Principal Component Analysis and Multiple Correspondence Analysis.

First we'll load the 'human' data frame and look at the structure and dimension of it.

```{r}
human <- read.csv(file = "./data/human.csv", row.names = 1)
str(human)
dim(human)
```
Here's a list of the variables included:

* **sec_edF2M:**
* **labour_F2M:**
* **eduexp:**
* **lifeexp:**
* **gni:**
* **matmort:**
* **ado_birth:**
* **rep_parl:**

Here's a graphical overview of the variables:

```{r}
library(GGally)
library(ggplot2)
p <- ggpairs(human, mapping = aes(alpha = 0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```

